Covariate analysis of MSSM reprocessed counts (with CQN normalisation and covariates adjusted)

### 2017-07-04 code/Rmd/MSSM_REPROCESSED.Rmd

Distribution of samples are: `r kable(tmp)`

### Conditional Quantile Normalisation (CQN)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E.NA = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts.NA = PROCESSED_COUNTS$filteredExprMatrix$counts 
PROCESSED_COUNTS$filteredExprMatrix$counts.NA[log.mat] = NA
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

CQN.GENE_EXPRESSION$E.NA[log.mat] = NA
CQN.GENE_EXPRESSION$E.zero = CQN.GENE_EXPRESSION$E.NA
CQN.GENE_EXPRESSION$E.zero[log.mat] = 0
```

```{r explore.data, cache=FALSE, fig.height=5, fig.width=10}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E.NA) %>%
  rownameToFirstColumn('Gene.ID') %>%
  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(BrainRegion+.~Diagnosis, scale = 'free')
p
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$E.zero))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Decomposition of expression
Clustering of library normalised unadjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=5, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E.zero, scale.=T, center = T)

@@

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E.zero)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E.zero, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

  # Calculate correlation values of random effects
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E.NA
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$individualIdentifier)

@@

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E.NA,
                            design = DM1$design,
                            weights = VOOM.GENE_EXPRESSION$weights,
                            block = COVARIATES$individualIdentifier, 
                            correlation = correlation$cor)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = list()
  RESIDUAL.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                        CQN.GENE_EXPRESSION$E.NA)
  RESIDUAL.GENE_EXPRESSION$expr.zero = RESIDUAL.GENE_EXPRESSION$expr.NA
  RESIDUAL.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.GENE_EXPRESSION$expr.zero)] = 0

@@

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  tmp.expr = RESIDUAL.GENE_EXPRESSION$expr.zero
  tmp.expr = tmp.expr[apply(tmp.expr, 1, var) != 0,]
  residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

@@

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residual.adj, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
tmp.expr = RESIDUAL.GENE_EXPRESSION$expr.zero
tmp.expr = tmp.expr[apply(tmp.expr, 1, var) != 0,]
residualSigCovars = runPCAandPlotCorrelations(tmp.expr, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

residualSigCovars[["PC_res"]][[2]]$plotData
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION$expr.zero))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2.1, fig.height=6, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION$expr.zero, scale.=T, center = T)

@@

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION$expr.zero)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)

@@

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
FIT = lmFit(VOOM.WEIGHTS)

@@

# Get residuals from differential expression model
RESIDUAL.SVA.GENE_EXPRESSION = list()
RESIDUAL.SVA.GENE_EXPRESSION$expr.NA = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E.NA)
RESIDUAL.SVA.GENE_EXPRESSION$expr.zero = RESIDUAL.SVA.GENE_EXPRESSION$expr.NA
RESIDUAL.SVA.GENE_EXPRESSION$expr.zero[is.na(RESIDUAL.SVA.GENE_EXPRESSION$expr.NA)] = 0

# Compute actual variance of all principal components
tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION$expr.zero)
actual.var = tmp$d^2/sum(tmp$d^2)

# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION$expr.zero, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])

@@

# Estimate surrogate variables
SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$E.zero,
                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv

@@

# Calculate correlation values of random effects
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$individualIdentifier)

@@

# Fit linear model using new weights and new design
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$E.NA)

@@

# Calculate correlation values of random effects
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$individualIdentifier)

@@

# Fit linear model using new weights and new design
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E.NA
VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)

# Residuals after normalisation
RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
                                                  CQN.GENE_EXPRESSION$E.NA)
```

@@

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts.NA %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
tmp = CQN.GENE_EXPRESSION$y
tmp[log.mat] = NA
tmp %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
tmp = CQN.GENE_EXPRESSION$offset
tmp[log.mat] = NA
tmp %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('MSSM_FP_STG_PHG_IFG_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

>>>>

Distribution of samples are: `r kable(tmp)`
### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t

# Back calculate counts for further analysis
LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
```

### Conditional Quantile Normalisation (CQN)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t
CQN.GENE_EXPRESSION$E = LOG.CPM

PROCESSED_COUNTS$filteredExprMatrix$counts = NEW.COUNTS
```

```{r explore.data, cache=FALSE, fig.height=5, fig.width=10}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E) %>%
  rownameToFirstColumn('Gene.ID') %>%
  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(BrainRegion+.~Diagnosis, scale = 'free')
p
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$E))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Decomposition of expression
Clustering of library normalised unadjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=5, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)

@@

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

  # Calculate correlation values of random effects
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$individualIdentifier)

@@

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E,
                            design = DM1$design,
                            weights = VOOM.GENE_EXPRESSION$weights,
                            block = COVARIATES$individualIdentifier, 
                            correlation = correlation$cor)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                CQN.GENE_EXPRESSION$E)

@@

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

@@

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
  print(loopCount)
}
RESIDUAL.GENE_EXPRESSION
modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residual.adj, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

residualSigCovars[["PC_res"]][[2]]$plotData
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2.1, fig.height=6, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

@@

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

# Calculate correlation values of random effects
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
correlation = parallelDuplicateCorrelation(VOOM.WEIGHTS, block = COVARIATES$individualIdentifier)

@@

# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS)

@@

# Get residuals from differential expression model
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E)

# Compute actual variance of all principal components
tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
actual.var = tmp$d^2/sum(tmp$d^2)

# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])

@@

# Estimate surrogate variables
SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$E,
                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv

@@

# Calculate correlation values of random effects
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
correlation = parallelDuplicateCorrelation(VOOM.SVA.WEIGHTS, block = COVARIATES$individualIdentifier)

@@

# Fit linear model using new weights and new design
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$E)

@@

# Calculate correlation values of random effects
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E
correlation = parallelDuplicateCorrelation(VOOM.NET.WEIGHTS, block = COVARIATES$individualIdentifier)

@@

# Fit linear model using new weights and new design
VOOM.NET.WEIGHTS$E = CQN.GENE_EXPRESSION$E
VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)

# Residuals after normalisation
RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
                                                  CQN.GENE_EXPRESSION$E)
```

@@

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('MSSM_FP_STG_PHG_IFG_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
